---
title: "Geoinside: Points in polygons via sf"
description: |
  Lets say we have some spatial areas (polygons) and then some spatial data points. And we are interested to find the area that each point falls into. An example could be finding the ICES statistical rectangles that a set of fishing operations fall under. This problem has been solved many times, this post is only a memo to oneself how this is done using functions in the sf-package using the tidyverse approach. The "geoinside" in the title is a reference to a function in the geo-packages, where such a solution has been available for over two decades, at least to some "insiders".
author:
  - name: Einar Hjörleifsson
    url: {https://heima.hafro.is/~einarhj/splatter/posts/2019-01-17-geoinside-points-in-polygon.html}
date: 2019-01-17
output:
  distill::distill_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Needed libraries:

```{r}
library(sf)
library(tidyverse)
```

## A generic case from scratch

### The code in short

Lets start with from scatch creating some 4 rectangle polygons and then a bunch of coordinates representing e.g. fishing haul location which we want to "assign" to each of the four rectangles. More details are provided below, but the final code-flow would be something like:

```{r}
# Generate the 4 rectangles:
icesr <- 
  data_frame(Rectangle = c(rep("58C2", 5), rep("58C1", 5), rep("57C2", 5), rep("57C1", 5)),
             lon = c( -28,  -28,  -27,  -27, -28,
                      -29,  -29,  -28,  -28, -27,
                      -28,  -28,  -27,  -27, -28,
                      -29,  -29,  -28,  -28, -27),
             lat = c(64.5, 65.0, 65.0, 64.5, 64.5,
                     64.5, 65.0, 65.0, 64.5, 64.5,
                     64.0, 64.5, 64.5, 64.0, 64.0,
                     64.0, 64.5, 64.5, 64.0, 64.0)) %>% 
  # Generate simple feature POINTS
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326) %>% 
  # Convert to sf MULTIPOINTS, "conditional" on variable Rectangle 
  group_by(Rectangle) %>% 
  summarise(do_union = FALSE) %>% 
  # Convert MULTIPOINTS to POLYGON
  st_cast("POLYGON")

# Generate some random (fishing haul) location
n <- 100
set.seed(314)
haul_location <- 
  data_frame(tow = 1:n,
             lon = runif(n, -29.1, -26.9),
             lat = runif(n,  63.9,  65.1)) %>% 
  # here want to keep the lon and the lat as attributes
  mutate(x = lon,
         y = lat) %>% 
  st_as_sf(coords = c("x", "y"),
           crs = 4326)

# Spatial joining
haul_location <- 
  haul_location %>% 
  st_join(icesr["Rectangle"])

# Visualize
ggplot() +
  geom_point(data = haul_location, aes(lon, lat, colour = Rectangle)) +
  geom_sf(data = icesr, alpha = 0.2, aes(fill = Rectangle)) +
  scale_fill_discrete(guide = FALSE) +
  labs(x = NULL, y = NULL)

# The final haul data:
haul_location
```

I.e. we within the haul dataframe the added variable "Rectangle" that each haul falls under. Hauls that are not within any of the four rectangles will have a missing (NA) rectangle variable name.

## Under the hood

If you want to know a little bit about the devil in detail, read on. If not skip the remainder of the post, it is really just a repeat of the stuff above.

### Generation of some polygons

Lets generate some data that we want in the **end** to represent some four spatial rectangle polygons (here specifially some four [ICES statistical rectangles](https://www.ices.dk/marine-data/maps/Pages/ICES-statistical-rectangles.aspx)):

```{r}
icesr <- 
  data_frame(Rectangle = c(rep("58C2", 5), rep("58C1", 5), rep("57C2", 5), rep("57C1", 5)),
             lon = c( -28,  -28,  -27,  -27, -28,
                      -29,  -29,  -28,  -28, -27,
                      -28,  -28,  -27,  -27, -28,
                      -29,  -29,  -28,  -28, -27),
             lat = c(64.5, 65.0, 65.0, 64.5, 64.5,
                     64.5, 65.0, 65.0, 64.5, 64.5,
                     64.0, 64.5, 64.5, 64.0, 64.0,
                     64.0, 64.5, 64.5, 64.0, 64.0))
```

What we have is a data frame of 20 records with 5 coordinates per rectangle which represent the boundary of each rectangle.

```{r}
icesr
```

The reason we have five points for each rectangle is that in order to create a proper spatial polygon (which could be of very irregular shape for that matter) one needs to "close the loop" by having the last coordinate the same as the first.

Before proceeding lets get a visual of the above via ggplot:

```{r}
ggplot() +
  geom_polygon(data = icesr, aes(lon, lat, fill = Rectangle))
```

Now to turn an R dataframe with coordinates into a spatial dataframe we need to specify what variables are the spatial coordinates and specify the coordinate reference system, this latter point has a lot of [devil in the details](https://en.wikipedia.org/wiki/Coordinate_system) which I suggest you ignore for now. Within the [tidyverse-sf](...) framework this can be achieved by using the sf-function 'st_as_sf' that converts an ordinary dataframe into a spatial simple-feature dataframe:

```{r}
icesr <- 
  icesr %>% 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326)
icesr
```

Basically we have still an R dataframe with 20 records, but the variable lon(gitude) and lat(itude) has been "move into" a single variable called geometry, that falls under the standard of [simple feature](https://en.wikipedia.org/wiki/Simple_Features). In this case "POINT". And then we have a bunch of "overhead" on the top. The whole "thing" is a bit of an add dataframe for anybody new to the sf-game, but those that want to dig deeper I highly recomend two free textbooks in different state of making:

1. [Geocomputation with R](https://geocompr.robinlovelace.net/)
2. [Spatial Data Science](https://keen-swartz-3146c4.netlify.com)

The next step is to convert each of the five sf-POINTS defining each statistical rectangle into four separate sf-POLYGON. Within the tidyverse environment we first aggregate the five points of each rectangle:

```{r}
icesr <- 
  icesr %>%  
  group_by(Rectangle) %>%
  summarise(do_union = FALSE)
icesr
```

Hmmm, this is a strange brew. The bottom line is that we have converted a dataframe of 20 records to only four records, one for each (ICES statistical) rectangle, with the five coordinates of each rectangle bundled into a single MULTIPOINT element.

In the final step we turn the multipoint to polygon using the 'st-cast' function:

```{r}
icesr <- 
  icesr %>% 
  st_cast("POLYGON")
icesr
```

Visually we have:

```{r}
ggplot() +
  geom_sf(data = icesr, aes(fill = Rectangle))
```

### Generating some random spatial data

Here we basically generate a set of some random spatial points conditional on spanning a slightly larger spatial range than the four rectangles generated above.

```{r}
n <- 100
set.seed(314)
haul_location <- 
  data_frame(tow = 1:n,
             lon = runif(n, -29.1, -26.9),
             lat = runif(n,  63.9,  65.1)) %>% 
  mutate(x = lon,
         y = lat) %>% 
  st_as_sf(coords = c("x", "y"),
           crs = 4326)
haul_location
```

### Spatial joining

Those familiar with the tidyverse lingo have probably used the ***_join sweep of functions that combines separate data frames based on some shared 'linking' variable(s), usually resulting one of the dataframe containing additional variable from the other dataframe. Spatial joining involves the same concept but uses the spatial data as the 'linking' variable.

In our case we are interested in adding the variable "Rectangle" from the icesr-dataframe to the haul-dataframe, where spatial object in the former is a polygon while in the latter we have spatial points. This final step is achieved by using the st_join function:

```{r}
haul_location <- 
  haul_location %>% 
  st_join(icesr["Rectangle"])
haul_location
```

