---
title: "PostgreSQL/PostGIS installation and then some"
description: |
  Again this blogpost is mostly a note to oneself. On installing PostgreSQL and its extention PostGIS on Fedora. And then loading and retrieving some spatial data via R.
author:
  - name: Einar Hjörleifsson
    url: https://splatter.netlify.com/posts/2020-03-16-matching-nearest-date.html
date: 05-29-2021
output:
  distill::distill_article:
    self_contained: false
draft: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE)
```

Since I had earlier installed postgres the first step, in order to ensure reproducability I delete the earlier installation (this is of course done in a terminal window):

```
sudo yum remove postgres\*
sudo rm -rf /var/lib/pgsql
```

The installation is then done by:

```
sudo dnf install postgis postgresql-server postgresql
sudo postgresql-setup --initdb
sudo systemctl start postgresql
```

Once things are installed one is ready for setting things up, like creating user name and database:

```
sudo su - postgres
createuser einar
createdb test_db -O einar
```

Next step was to grant privileges for the user:

```
psql -d test_db -c "grant all privileges on database test_db to einar;"
```

The above should be sufficient if one is only working with non-spatial data. But to include the PostGIS one needs to add that extention. This has to be done for each database that one generates:

```
psql -d test_db -c "CREATE EXTENSION postgis;"
psql -d test_db -c "CREATE EXTENSION postgis_topology;"
```

Once in place we are now ready to switch to R. Here we first obtain some spatial data:

```{r}
library(tidyverse)
library(sf)
bc <- 
  read_sf("ftp://ftp.hafro.is/pub/data/shapes/bormicon.gpkg")
```

The data has a geometry that are polygons:

```{r}
bc %>% glimpse()
```

Visually we have:

```{r}
bc %>% 
  ggplot() + 
  geom_sf(aes(fill = name)) +
  theme(legend.position = "none")
```

We are now ready to load the data into the database. We first need to setup the connetion and then we will simply use `st_write` to upload the data to the database:

```{r}
library(RPostgreSQL)
con <- dbConnect(drv = "PostgreSQL", dbname = "test_db")
bc %>% 
  st_write(dsn = con, layer = "bormicon")
rm(bc)  # nothing up my sleave
```

Now we could try to retrieve that data via a usual `tbl`-function:

```{r}
bc <- tbl(con, "bormicon")
bc %>% glimpse()
```

Here we get a warning message and the geometry looks very unfamiliar (is in binary format). It is still of the class expected and one can take a peek at the sql-query:

```{r}
bc %>% class()
bc %>% show_query()
```

Now one could do the usual database acrobatics via {dbplyr} on this but only in the non-spatial form. To get the proper spatial format one simply uses the `st_read`-function:

```{r}
rm(bc)  # nothing up my sleave
bc <- st_read(con, layer = "bormicon")
bc %>% class()
```

The class here is however a spatial tibble that is fully loaded into R. I.e. we can not send any spatial sf-verbs to the database. There are ways to generate spatial queries to the database via sql-construction but there is as of yet no spatial backend that translates sf-function to sql. But that could be coming soon, check out e.g. [Database interoperability for spatial objects in R](https://etiennebr.github.io/spatial-database-proposal/out/isc-proposal.html).

```{r}
DBI::dbDisconnect(con)
```


