---
title: "Merging logbook and vms"
description: |
  We have two datasets: 1) Logbook data were each fishing haul is a record and has start and end time as well as geographical start and end coordinates. 2) VMS data were each record is spatial point in time.
  Ideally one should try to attempt to use both time and space to join the data, but here we will only use the time dimension. In essence the problem to solve is matching time (vms-interval problem to solve.

author:
  - name: Einar Hjörleifsson
    url: https://heima.hafro.is/~einarhj/splatter/posts/2020-02-29-merging-logbooks-and-vms.html
date: 02-29-2020
output:
  distill::distill_article:
    self_contained: false
draft: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(lubridate)
library(tidyverse)
library(sf)
library(mapdeck)
```

### The data

#### VMS data

The dataset are 8918 [AIS](https://en.wikipedia.org/wiki/Automatic_identification_system)/[VMS](https://en.wikipedia.org/wiki/Vessel_monitoring_system) records of four vessels that participated in the Icelandic spring survey in 2019. The variables are as follows:


```
| Option           | Description                                    |
|------------------|------------------------------------------------|
| `rows.print`     | Maximum rows to print per page.                |
| `max.print`      | Maximum rows in the table (defaults to 1000).  |
| `cols.print`     | Maximum columns in the table (defaults to 10). |
| `rownames.print` | Print row names as part of the table.          |
```

* **vid**: Vessel identification number
* **vessel**: Vessel name
* **time**: Time in UTC
* **lon**: Longitude in decimal degrees
* **lat**: Latitude in decimal degrees
* **speed**: Vessel speed in knots
* **heading**: Vessel headings in degrees

**Importing the data into R**:

```{r}
d <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb_vms2019.csv")
```


The example datasets are the Icelandic spring groundfish survey from 2019, where four different vessels participated.

Lets use the Icelandic survey data as our logbook data and the vms-data from the vessels that participated in the 2019 survey.

```{r}
vms <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb_vms2019.csv") %>% 
  arrange(vid, time)
lgs <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb.csv") %>% 
  filter(year == 2019) %>% 
  select(id, vid, t1, t2, lon1, lat1, lon2, lat2)
```





### In space

Lets first look at the spatial dimensions, here zooming into an area:


```{r echo = FALSE}
key <- "pk.eyJ1IjoiZmlzaHZpY2UiLCJhIjoiY2p0MXQ5dThpMDZqeDQ5bHM0MGx4dHI5cyJ9.Fed_z7mv_TgTWDRjiavU3A"
set_token(key)
```

```{r}
mapdeck() %>% 
  add_scatterplot(data = vms,
                  lon = "lon", 
                  lat = "lat",
                  fill_colour = "speed",
                  palette = "inferno")
```



```{r, fig.height=4}
ggplot() +
  theme_bw() +
  geom_segment(data = lgs,
               aes(lon1, lat1,
                   xend = lon2, yend = lat2),
               colour = "grey",
               lwd = 2) +
  geom_point(data = vms,
             aes(lon, lat, colour = speed),
             size = 2) +
  scale_color_viridis_c(option = "B", direction = -1) +
  geom_path(data = vms,
            aes(lon, lat, group = vid),
            colour = "grey") +
  coord_quickmap(xlim = c(-20, -19.25),
                 ylim = c(66.7, 66.85)) +
  labs(x = NULL, y = NULL)
```

The plot shows:

* The thick "grey" line is a line represents the shortest path between start and end position of the tow.
* The points represents the ping observations, and the colour the vessel speed.
* The thin grey line connects the points between pings.

On a fine space-scale we observe that the spatial points in the vms and start and end points of a tow do not intersect.

### By time

Both datasets also have a time dimension, here we zoom into one day of the survey:

```{r}
ggplot() +
  theme_bw() +
  geom_rect(data = lgs,
            aes(xmin = t1, xmax = t2),
            ymin = -Inf, ymax = Inf,
            fill = "grey") + 
  geom_point(data = vms, aes(time, speed), size = 1) +
  facet_grid(vid ~ .) +
  scale_x_datetime(limits = c(ymd_hms("2019-03-13 00:00:00"),
                              ymd_hms("2019-03-14 00:00:00")))
```

The plot shows:

* The grey area is the time between the recorded start (t1) and end (t2) of the tow.
* The points represents the vms-ping observations, speed being indicated by the y-axis.

On a finer time-scale we observe that the time in the vms and the time in the do not intersect.

### More observation on the vms data

```{r}
ggplot() +
  geom_histogram(data = vms,
                 aes(speed))
vms %>% 
  group_by(vid) %>% 
  mutate(hz = time - lag(time),
         hz = as.numeric(hz) / 60) %>% 
  ggplot() +
  geom_histogram(aes(hz)) +
  scale_x_continuous(name = "Ping frequency [mins]",
                     lim = c(0, 15))
```


## Solving the problem using time

Lets just take **one vessel** for now, the thinking being that at a later time one can scale the code up to deal with all the vessels. The steps we will take are:

1. Transform the logbook data such time is stored as a single variable, using an additional variable (startend) to indicate the if the time value refers to the start or end of the haul.
2. Round the vms data to the nearest minute
3. Create a dataframe with time at **one** minute interval, going from the minimum and to the maximum time in the vms data.
4. Join this dataframe with the vms-data
5. Interpolate the geographical positions as well as speed and heading (an option)
6. Bind the frame from step 5 with the logbook data and arrange by time
7. Do some coding acrobatics to assign the tow id (from the logbooks) to vms data
8. Interpolate again, filling in values for the logbook records

### The code 

```{r}
VID <- 1131

# Step 1
smb2 <- 
  lgs %>% 
  filter(vid == VID) %>% 
  select(vid, id, t1, t2) %>%
  pivot_longer(cols = c(t1, t2),
               names_to = "startend",
               values_to = "time") %>% 
  arrange(vid, time) %>%
  mutate(year = year(time))

# Step 2
vms2 <- 
  vms %>% 
  filter(vid == VID) %>% 
  # make sure we have unique records
  distinct() %>% 
  mutate(time = round_date(time, "minutes"))


time <- 
  seq(from = min(vms2$time),
      to   = max(vms2$time),
      by   = "1 min")
g <- 
  # Step 3, using vector created above
  tibble(time = time,
         vid = VID) %>% 
  # Step 4
  left_join(vms2 %>% mutate(meas = TRUE),
            by = c("time", "vid")) %>% 
  # Step 5
  group_by(vid) %>% 
  mutate(y = 1:n()) %>% 
  mutate(lon = approx(y, lon, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         lat = approx(y, lat, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         speed = approx(y, speed, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         heading = approx(y, heading, y, method = "linear", rule = 1, f = 0, ties = mean)$y) %>%
  # Step 6
  bind_rows(smb2) %>% 
  arrange(vid, time) %>%
  # Step 7
  mutate(x = case_when(startend == "t1" ~ 1,
                       startend == "t2" ~ -1,
                       TRUE ~ 0)) %>%
  mutate(x = cumsum(x)) %>%
  # fill "does too much", ...
  fill(id) %>%
  # hence we need do this:
  mutate(id = ifelse(x == 1 | startend == "t2", id, NA_integer_)) %>% 
  # Step 8
  mutate(y = 1:n(),
         lon = approx(y, lon, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         lat = approx(y, lat, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         speed = approx(y, speed, y, method = "linear", rule = 1, f = 0, ties = mean)$y,
         heading = approx(y, heading, y, method = "linear", rule = 1, f = 0, ties = mean)$y) %>% 
  mutate(fishing = ifelse(!is.na(id), TRUE, FALSE))
```

### The results

Lets take a visual peek at what we have done:

```{r, fig.height=4}
ggplot() +
  theme_bw() +
  geom_segment(data = lgs %>% filter(vid == VID),
               aes(lon1, lat1,
                   xend = lon2, yend = lat2),
               colour = "grey",
               lwd = 2) +
  geom_point(data = g,
             aes(lon, lat, colour = fishing),
             size = 0.4) +
  scale_color_brewer(palette = "Set1") +
  coord_quickmap(xlim = c(-20, -19.25),
                 ylim = c(66.7, 66.85)) +
  labs(x = NULL, y = NULL)
```

```{r}
ggplot() +
  theme_bw() +
  geom_rect(data = lgs %>% filter(vid == VID),
            aes(xmin = t1, xmax = t2),
            ymin = -Inf, ymax = Inf,
            fill = "grey") + 
  geom_point(data = g, aes(time, speed, colour = fishing), size = 1) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(vid ~ .) +
  scale_x_datetime(limits = c(ymd_hms("2019-03-13 00:00:00"),
                              ymd_hms("2019-03-14 00:00:00")))
```
